---
title: "ps1.qmd"
author: "Caleb Masshardt"
format: pdf
editor: visual
---

## Gov 51 PSet 1

Caleb Masshardt

February 11, 2026

GitHub Repository URL: <https://github.com/cmasshardt/Gov-51-PSet-1.git>

## 1. Project Setup (10 points)

Instructions from problem set followed. See GitHub repository URL above.

## 2. Get to Know Your Data (15 points)

*1. Load the tidyverse and read in your ACS data.*

```{r}
library(tidyverse)

acs <- read.csv("./data/raw/acs2024.csv")
```

*2. Look up what each variable in your dataset means.*

TRANTIME: Commute time (minutes) of individual. A value of "0" indicates that the person has no commute time (either unemployed, not in the labor market, or works from home).

PERWT: How many persons in the population are represented by A given person in the IMUPS sample.

*3. Create new variables from the raw IMUPS code.*

```{r}
#Check range of SEX to see if there are NA values (labeled as 9)
range(acs$SEX)
```

```{r}
#Female variable: equals 1 if female, 0 otherwise female
acs <- acs %>%
  mutate(female = SEX-1)

```

```{r}
# Check range of EDUC to see if there are NA values (labeled as 99)
range(acs$EDUC)
```

```{r}
#Education dummy variables
#less_than_hs
acs <- acs %>%
  mutate(less_than_hs = as.numeric(EDUC==c(1,2)))
#hs_only
acs <- acs %>%
  mutate(hs_only = as.numeric(EDUC==c(3,4,5,6)))
#some_college
acs <- acs %>%
  mutate(some_college = as.numeric(EDUC==c(7,8,9)))
#college_only
acs <- acs %>%
  mutate(college_only = as.numeric(EDUC==10))
#advanced_degree
acs <- acs %>%
  mutate(advanced_degree = as.numeric(EDUC==11))
```

```{r}
#Check range of EMPSTAT to see if there are NA values (labeled as 9)
range(acs$EMPSTAT)
```

```{r}
#Employment dummy variables
#employed
acs <- acs %>%
  mutate(employed = as.numeric(EMPSTAT==1))
#unemployed
acs <- acs %>%
  mutate(unemployed = as.numeric(EMPSTAT==2))
#not_in_labor_force
acs <- acs %>%
  mutate(not_in_labor_force = as.numeric(EMPSTAT==3))
```

Remove N/A values (expressed as 999999) from INCTOT
```{r}

```


*4. Why does this matter? In 2–3 sentences, explain what could go wrong if you accidentally treated “N/A” codes as real zeros when calculating summary statistics. Give a concrete example using one of the variables in your dataset.*

Since the N/A values for the variables in this dataset are represented by numerals accidentally treating these representations of N/A values as real zeros would lead to very different values for the mean, standard deviation, and maximum, which are vulnerable to strong influence from outliers. For example, for the variable AGE, N/A values are written as 999. If one of the observations in the dataset for age were N/A (999), the summary statistics for age would be strongly and inaccurately influenced by the enormous value of 999. As such, the mean, standard deviation, and maximum would all be incorrect.

*5. Create a summary statistics table for your dataset.*

```{r}
#Creating summary statistics table
summary_stats <- data.frame(
  Variable = c("Age", "Female", "Less than High School", "High School Only", "Some College", "College Only", "Advanced Degree", "Employed", "Unemployed", "Not in Labor Force", "Commute Time (mins)", "Total Income ($)"),
  N = c(length(acs$AGE),
        length(acs$female), length(acs$less_than_hs), length(acs$hs_only),
        length(acs$some_college), length(acs$college_only),
        length(acs$advanced_degree), length(acs$employed),
        length(acs$unemployed), length(acs$not_in_labor_force),
        length(acs$TRANTIME), length(acs$INCTOT)),
  Mean = c(mean(acs$AGE),
        mean(acs$female), mean(acs$less_than_hs), mean(acs$hs_only),
        mean(acs$some_college), mean(acs$college_only),
        mean(acs$advanced_degree), mean(acs$employed),
        mean(acs$unemployed), mean(acs$not_in_labor_force),
        mean(acs$TRANTIME), mean(acs$INCTOT)),
  SD = c(sd(acs$AGE),
        sd(acs$female), sd(acs$less_than_hs), sd(acs$hs_only),
        sd(acs$some_college), sd(acs$college_only),
        sd(acs$advanced_degree), sd(acs$employed),
        sd(acs$unemployed), sd(acs$not_in_labor_force),
        sd(acs$TRANTIME), sd(acs$INCTOT)),
  Min = c(min(acs$AGE),
        min(acs$female), min(acs$less_than_hs), min(acs$hs_only),
        min(acs$some_college), min(acs$college_only),
        min(acs$advanced_degree), min(acs$employed),
        min(acs$unemployed), min(acs$not_in_labor_force),
        min(acs$TRANTIME), min(acs$INCTOT)),
  Max = c(max(acs$AGE),
        max(acs$female), max(acs$less_than_hs), max(acs$hs_only),
        max(acs$some_college), max(acs$college_only),
        max(acs$advanced_degree), max(acs$employed),
        max(acs$unemployed), max(acs$not_in_labor_force),
        max(acs$TRANTIME), max(acs$INCTOT))
)

#Round the stats
stats_rounded <- summary_stats
stats_rounded[, -1] <- round(summary_stats[, -1], 2)
stats_rounded

#For table presentation
library(knitr)

stats_rounded |>
  kable(
    col.names = c("Variable", "N", "Mean", "Std Dev", "Min", "Max") ,
    digits = 2 ,
    caption = "Summary Statistics for 2024 ACS Sample"
  )
```
